---
title: "DATA101 Assignment 1, Semester 2 2025"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

**Due Date:** 25 July 2025

**Student ID Number:** 83972706

**Total Marks:** 20

**Instructions:**

-   Complete all questions in R Markdown format

-   If using this .Rmd file for your solution, please enter your student ID number in the space provided above

-   Submit your completed assignment as a PDF through LEARN by 11:59pm on the due date

-   Show all your R code and explain your reasoning where required

-   Code that produces errors will receive minimal marks

## Question One (7 marks)

Below you are given the daily step counts recorded by a fitness tracker for 10 consecutive days:

\begin{tabular}{ccccc cccccc} 8453 & 6721 & 9546 & 7832 & 5298 & 10234 & 8765 & 7123 & 9871 & 6540 \end{tabular}

Answer the following:

a)  Write the R code to create a vector with these daily step counts that allows you to perform statistical analysis on them. (2 marks)

```{r}
vector <- c(8453, 6721, 9546, 7832, 5298, 10234, 8765, 7123, 9871, 6540)
vector
```

b)  Calculate the mean daily steps and the standard deviation of these measurements using R. (1 mark)

```{r}
vector_mean <- mean(vector)
vector_std <- sd(vector)
vector_mean
vector_std
```

c)  What is the median daily step count? Explain the steps you would follow to calculate this median manually without using R functions. (2 marks)

```{r}
vector_median <- median(vector)
vector_median
```

To do without R functions I would arrange the values in ascending order, if the number of values is odd then the middle value is the median, if it is even then it would be the average of the two middle values

d)  Create a histogram and a boxplot of these step counts. Based on these visualizations, comment on the distribution of the data and identify any potential outliers. (2 marks)

## Create a histogram of the vector

```{r}
vector <- c(8453, 6721, 9546, 7832, 5298, 10234, 8765, 7123, 9871, 6540)
hist(vector)
```

## Create a box plot of the vector

```{r}
vector <- c(8453, 6721, 9546, 7832, 5298, 10234, 8765, 7123, 9871, 6540)
boxplot(vector, main = "Boxplot of Values", ylab = "Values", col = "lightblue")
```

The shape of the data appears to be quite symmetric.

The data ranges from just under 5,300 to over 10,200, indicating a wide spread.

There are no obvious outliers

## Question Two: (6 marks)

The dataset `customer_data_2025.csv` contains information about customers of an online retailer, including their `time_on_site` (in minutes), `purchase_amount` (in dollars), `age` (in years), `clicks_before_purchase` (number of clicks), and `return_customer` (yes/no). The dataset can be downloaded from LEARN.

a)  In this scenario, identify which variables could be considered explanatory variables and which would be the response variable if we wanted to predict purchase amount. Justify your answer. (1 mark)

### Explanatory:

- time_on_site
- clicks_before_purchase
- age

time_on_site and clicks_before_purchase are both included as explanatory variables because they represent meaningful aspects of customer engagement. More time or more clicks may reflect greater interest or deliberation, which could reasonably influence how much a customer ends up spending (purchase_amount).

### Response

- purchase_amount

This variable will reflect how much the customer spent on the site so if this is lower we can see the customer spent less money on the site

b)  Create a scatterplot to visualize the relationship between time spent on the website and purchase amount. Add appropriate axis labels and a title to your plot. Include a color dimension to represent whether the customer is a return customer or not. (1 mark)

## Create a scatterplot on the data

```{r}
customer_data <- read.csv('DATA/DATA101_Assignment1_Data_customer_data_2025.csv')
Xs <- customer_data$time_on_sit # TODO
Ys <- customer_data$purchase_amount #TODO
plot(Xs, Ys, xlab = 'Time spent on site (Mins)', ylab = 'Purchase amount ($)', main = 'Time spent on site Vs Purchase amount')

```

c)  Transform the data by creating new variables: `log_time` = `log(time_on_site)` and `log_purchase` = `log(purchase_amount+1)`. Generate a scatterplot of these transformed variables. Additionally, create another visualization showing the relationship between `clicks_before_purchase` and `purchase_amount`, grouped by age categories (e.g., `18 - 30, 31 - 45, 46 - 60, 61+`). Explain how these transformations and groupings affect your understanding of the relationships in the data. (2 marks)

```{r}
customer_data <- read.csv('DATA/DATA101_Assignment1_Data_customer_data_2025.csv')
log_time <- log(customer_data$time_on_site)
log_purchase <- log(customer_data$purchase_amount+1)
plot(log_time, log_purchase, xlab = 'Time spent on site (Mins)', ylab = 'Purchase amount ($)', main = 'Time spent on site Vs Purchase amount (Log transformed)')
```

# Create a color for each age group

```{r}
customer_data <- read.csv('DATA/DATA101_Assignment1_Data_customer_data_2025.csv')
age <- customer_data$age
age_group <- cut(
  age,
  breaks = c(18, 30, 45, 60, Inf),
  labels = c("18-30", "31-45", "46-60", "61+"),
  right = TRUE
)
clicks_before_purchase <- customer_data$clicks_before_purchase
purchase_amount <- customer_data$purchase_amount
age_colors <- as.numeric(age_group)
plot(
  clicks_before_purchase, purchase_amount,
  col = age_colors,
  pch = 16,
  xlab = "Clicks Before Purchase",
  ylab = "Purchase Amount ($)",
  main = "Clicks vs Purchase Amount by Age Group"
)

legend("topright", legend = levels(age_group), pch=16, col = 1:4)
```

This allows me to see that the amount of clicks a user does on the site is reflected in their purchase amount but it appears that this is only the case when the user is \> 61 years of age.

d)  After fitting a linear model to the transformed data, you find that:

```{r, echo=TRUE, eval=FALSE}
purchase_amount = 2.75 * (time_on_site)^0.65 * (age)^0.12 * (0.95)^(clicks_before_purchase - 1)
```

Use this model to predict the purchase amount for a $35$ year old return customer who spends $25$ minutes on the website and makes $12$ clicks before purchasing. (1 mark)

```{r}
purchase_amount = 2.75 * (25^0.65) * (35^0.12) * (0.95^11)
```

e)  What additional information should a data scientist include when communicating the above result to the company? Consider aspects like prediction intervals, model assumptions, and limitations of the analysis. (1 mark)

They should include a prediction interval to show uncertainty, explain model assumptions (e.g. variable relationships, independence), and note limitations like potential data bias or changes in customer behavior that may affect accuracy.

## Question Three: (7 marks)

You have been asked to determine the proportion of Auckland City residents who use public transportation as their primary method of commuting to work. Answer the following:

a)  What is the population of interest? (1 mark)

Auckland city residents that commute to work

b)  As it is not feasible to interview all residents, you are asked to draw a sample of $1500$ residents from the city's database. To minimise bias, what property should the sampling procedure have? (1 mark)

The sample should be randomly selected, so that it is representative of the population and reduces selection bias

c)  Are you trying to describe cause and effect, that is conduct an analytic study, or just describing, that is conduct a purely descriptive study? (1 mark)

Purely descriptive study

d)  What is the population parameter in this example? How would you obtain an estimate of the population parameter in this example? (2 marks)

1550 (The size of the sample I was asked to collect) (Number of people that use transport as their primary commute)/1500

e)  If you are undertaking this study now, i.e. in the year 2025, describe two potential biases that could arise in your study. (2 marks)

<!-- -->

1.  People who use public transport might be more willing to participate in surveys about commuting, skewing the results.

2.  The proportion of people commuting at all may have changed significantly after the COVID-19 lockdowns, which could distort the apparent use of public transport among "commuters".
